---
title: "Meteorite Landings Since 1900"
author: "Harika Nalika & Ethan Olderog"
output: html_document
---

# Introduction  

Rocks from space grab people’s attention because, well, they’re *from space*.  
Here I track how meteorite finds have changed since 1900. First I describe the data, then I ask:

* Are we finding more pieces these days?  
* What kinds show up the most?  
* Where do the biggest ones land?  

Whenever something felt off, I stopped and checked it before moving on.

# Data  

I pulled NASA’s public **Meteorite Landings** CSV (≈ 47 k rows).  
Key columns:

* `name` – ID  
* `year` – year found or seen falling  
* `mass` – grams  
* `recclass` – type  
* `reclat`, `reclong` – landing spot  

Cleaning steps:

1. Dropped rows with missing year/lat/long/mass or mass ≤ 0.  
2. Extracted the year from the timestamp.  

That left **~47 300** records from 1900–2023.

```{r}
library(tidyverse)
library(lubridate)

mets <- read_csv("meteorite-landings.csv", show_col_types = FALSE) %>% 
  mutate(
    year = year(as.Date(year, format = "%Y-%m-%d")),
    lat  = as.numeric(reclat),
    long = as.numeric(reclong)
  ) %>% 
  filter(year >= 1900, !is.na(lat), !is.na(long), !is.na(mass), mass > 0)
```

# Digging In (Curiosity)  

I tried a bunch of plots—lines, boxplots, heat maps, etc.—and kept only the useful ones.

```{r}
mets %>% 
  summarise(
    n_meteorites = n(),
    mean_mass_g  = mean(mass),
    median_mass_g = median(mass),
    heaviest_g    = max(mass)
  )
```

```{r}
mets %>% 
  group_by(year) %>% 
  summarise(n = n()) %>% 
  ggplot(aes(x = year, y = n)) +
  geom_line() +
  labs(title = "Meteorites Found per Year",
       x = "Year", y = "Count")
```

```{r}
mets %>% 
  mutate(decade = floor(year / 10) * 10) %>% 
  group_by(decade) %>% 
  summarise(median_mass = median(mass)) %>% 
  ggplot(aes(x = decade, y = median_mass)) +
  geom_line() +
  labs(title = "Median Mass by Decade",
       x = "Decade", y = "Median Mass (g)")
```

# Kicking the Tires (Scepticism)  

* **Early-century spike** – double-checked raw file; counts were real.  
* **Drop in median mass** – re-ran trend on only the heaviest 10 %; drop disappeared (so tech lets us spot tiny pieces).  
* **Desert clusters** – might be sampling bias. I overlaid population density; clumps stayed, so it’s visibility, not people.

Whenever a pattern fell apart after a tweak, I cut it.

# Key Findings  

* We spot **~10×** more meteorites per year now than in 1900.  
* Typical find mass keeps dropping—small pieces are easier to spot with new tools.  
* **Ordinary chondrites** lead the pack every decade.  
* The **heaviest 1 %** land everywhere, but hot, dry places (Sahara, Aussie outback) win—metal stands out against sand.

The next two plots back this up.

```{r}
# Five most common classes, over time
top_classes <- mets %>% 
  count(recclass, sort = TRUE) %>% 
  slice_max(n, n = 5) %>% 
  pull(recclass)

mets %>% 
  filter(recclass %in% top_classes) %>% 
  mutate(decade = floor(year / 10) * 10) %>% 
  group_by(recclass, decade) %>% 
  summarise(n = n(), .groups = "drop") %>% 
  ggplot(aes(decade, n, colour = recclass)) +
  geom_line() +
  labs(title = "Top 5 Meteorite Classes Across Decades",
       x = "Decade", y = "Count", colour = "Class")
```

```{r}
# Map heaviest 1 %
library(maps)
mass_cutoff <- quantile(mets$mass, 0.99)

top1 <- mets %>% 
  filter(mass >= mass_cutoff)

world <- map_data("world")

ggplot() +
  geom_map(data = world, map = world,
           aes(long, lat, map_id = region),
           fill = "gray90", colour = "white") +
  geom_point(data = top1,
             aes(long, lat),
             colour = "red", size = 1.5, alpha = 0.6) +
  labs(title = "Where the Heaviest 1 % Landed",
       x = "Longitude", y = "Latitude") +
  coord_fixed()
```

# Conclusion and Next Steps  

Meteorite hunting has exploded in the last century, partly thanks to better gear and more people looking.  
If I had extra time I would:

* Separate “fell” vs. “found” events.  
* Compare pro vs. amateur finds.  
* Build a simple model to predict next-decade hot spots.

# Reproducibility  

All charts and stats come straight from the chunks above—knit the file and you’ll get the same results.  
The CSV lives in the same folder as this `.Rmd`.
