---
title: "Meteorite Landings Since 1900"
author: "Harika Nalika & Ethan Olderog"
output:
  html_document:
    self_contained: true
    toc: false
  github_document: default
---
```{r setup, include=FALSE}
# Global chunk options
knitr::opts_chunk$set(
  echo = TRUE, message = FALSE, warning = FALSE,
  fig.width = 7, fig.height = 4.5
)

# Make sure ggplot2 is loaded before theme_set
if (!requireNamespace("ggplot2", quietly = TRUE)) {
  stop('Package "ggplot2" is required. Please install it with: install.packages("ggplot2")')
}
library(ggplot2)
ggplot2::theme_set(ggplot2::theme_minimal(base_size = 12))
```

# Introduction

Space rocks are interesting because they come from outside Earth.  
This project looks at meteorite records since 1900. I will answer three simple questions:

- Are we finding more pieces now than before?
- What kinds show up the most?
- Where do the biggest ones show up?

If something did not look right, I checked it before moving on.

# Data

I used NASA's public "Meteorite Landings" CSV.  
Important columns:

- `name` - an id for the piece
- `year` - the year it was found or seen falling
- `mass` - the mass in grams
- `recclass` - the type
- `reclat`, `reclong` - the location (or I can parse it from a text field named `GeoLocation` if needed)

Cleaning steps I used:

1. Drop rows with missing year, latitude, longitude, or mass, or with mass less than or equal to 0.
2. Turn date-time into just a year when needed.
3. Keep only years from 1900 up to the current year, so there are no future years in the plots.

```{r load-data}
# Libraries
for (pkg in c("readr","dplyr","tidyr","lubridate")) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop(paste('Package "', pkg, '" is required. install.packages("', pkg, '")', sep = ""))
  }
}
library(readr); library(dplyr); library(tidyr); library(lubridate)

# Require the CSV to be in the same folder; fail fast with a clear message
if (!file.exists("meteorite-landings.csv")) {
  stop('Could not find "meteorite-landings.csv" in the working directory. Place the CSV next to this Rmd and knit again.')
}

raw_mets <- readr::read_csv("meteorite-landings.csv", guess_max = 100000, show_col_types = FALSE)

# Some copies use "mass (g)" instead of "mass"
if (!("mass" %in% names(raw_mets)) && ("mass (g)" %in% names(raw_mets))) {
  raw_mets <- dplyr::rename(raw_mets, mass = `mass (g)`)
}

# If reclat and reclong are missing but GeoLocation exists like "(lat, long)",
# extract them using simple string steps (no backslash escapes)
if (!("reclat" %in% names(raw_mets)) && ("GeoLocation" %in% names(raw_mets))) {
  tmp <- raw_mets %>%
    mutate(GeoLocation = as.character(GeoLocation)) %>%
    mutate(GeoLocation_clean = gsub("[()]", "", GeoLocation)) %>%
    tidyr::separate(GeoLocation_clean, into = c("lat_extracted", "long_extracted"),
                    sep = ",", fill = "right", extra = "drop", remove = FALSE) %>%
    mutate(lat_extracted = trimws(lat_extracted),
           long_extracted = trimws(long_extracted))
  if ("lat_extracted" %in% names(tmp) && "long_extracted" %in% names(tmp)) {
    raw_mets <- tmp %>%
      mutate(reclat = lat_extracted, reclong = long_extracted) %>%
      select(-lat_extracted, -long_extracted, -GeoLocation_clean)
  }
}

# Helper: safe numeric conversion
to_numeric <- function(x) {
  if (is.numeric(x)) return(x)
  suppressWarnings(as.numeric(as.character(x)))
}

current_year <- lubridate::year(Sys.Date())

# Parse and clean
mets <- raw_mets %>%
  mutate(
    year_raw = year,
    # Try several date formats; fall back to integer if already just a year
    year = suppressWarnings(lubridate::year(lubridate::ymd_hms(as.character(year)))),
    year = ifelse(is.na(year), suppressWarnings(lubridate::year(lubridate::ymd(as.character(year)))), year),
    year = ifelse(is.na(year), suppressWarnings(as.integer(as.character(year_raw))), year),
    lat  = to_numeric(reclat),
    long = to_numeric(reclong),
    mass = to_numeric(mass)
  ) %>%
  filter(!is.na(year), year >= 1900, year <= current_year,
         !is.na(lat), !is.na(long),
         !is.na(mass), mass > 0) %>%
  mutate(decade = floor(year / 10) * 10)

# Quick check
dplyr::glimpse(mets)
```

# Results and simple explanations

I tried a few plots and kept the ones that answered the questions directly.

```{r summary-stats, fig.cap="Summary of the dataset after cleaning."}
summary_tbl <- mets %>%
  summarise(
    n_meteorites  = n(),
    mean_mass_g   = mean(mass, na.rm = TRUE),
    median_mass_g = median(mass, na.rm = TRUE),
    heaviest_g    = max(mass, na.rm = TRUE)
  )

knitr::kable(summary_tbl, caption = "Overall summary statistics")
```

**Plain explanation:**  
This table tells us how many rows we have and the basic mass numbers. If the mean is much bigger than the median, it means there are a few very heavy meteorites that pull up the average. That is normal with this kind of data.

```{r per-year-count, fig.cap="Meteorites found per year (no future years)."}
mets %>%
  count(year, name = "n") %>%
  ggplot(aes(x = year, y = n)) +
  geom_line(na.rm = TRUE) +
  labs(title = "Meteorites Found per Year",
       x = "Year", y = "Count")
```

**Plain explanation:**  
Each point shows how many meteorites were recorded in that year. The line goes up and down, but overall it rises. This means we record more finds now than in the early 1900s. Better tools and more searching likely help.

```{r median-mass-by-decade, fig.cap="Typical meteorite mass by decade (median)."}
mets %>%
  group_by(decade) %>%
  summarise(median_mass = median(mass, na.rm = TRUE), .groups = "drop") %>%
  ggplot(aes(x = decade, y = median_mass)) +
  geom_line(na.rm = TRUE) +
  labs(title = "Median Mass by Decade",
       x = "Decade", y = "Median Mass (g)")
```

**Plain explanation:**  
This shows the typical mass in each decade. If the line is lower in recent decades, that means we are logging more small pieces now. That makes sense because small fragments are easier to find with modern methods.

```{r top-classes-over-time, fig.cap="Top 5 meteorite classes across decades."}
# Top 5 classes across decades
class_counts <- mets %>%
  filter(!is.na(recclass), recclass != "") %>%
  count(recclass, sort = TRUE)

top_classes <- class_counts %>%
  slice_head(n = 5) %>%
  pull(recclass)

class_by_decade <- mets %>%
  filter(recclass %in% top_classes) %>%
  group_by(recclass, decade) %>%
  summarise(n = n(), .groups = "drop")

stopifnot(nrow(class_by_decade) > 0)

ggplot(class_by_decade, aes(decade, n, group = recclass)) +
  geom_line(aes(linetype = recclass)) +
  labs(title = "Top 5 Meteorite Classes Across Decades",
       x = "Decade", y = "Count", linetype = "Class")
```

**Plain explanation:**  
This compares the most common types over time. If a type stays high across many decades, it is consistently common. In public data, ordinary chondrites are usually the most common. You can see that pattern here too.

```{r map-heaviest, fig.cap="Where the heaviest 1 percent of meteorites were recorded."}
# Map heaviest 1 percent with a fallback if the maps package is not installed
mass_cutoff <- stats::quantile(mets$mass, 0.99, na.rm = TRUE)

top1 <- mets %>%
  filter(!is.na(lat), !is.na(long), mass >= mass_cutoff)

if (nrow(top1) == 0) {
  warning("Heaviest 1 percent subset is empty. Check mass parsing.")
}

if (requireNamespace("maps", quietly = TRUE)) {
  world <- ggplot2::map_data("world")

  ggplot() +
    geom_polygon(data = world,
                 mapping = aes(long, lat, group = group),
                 fill = "gray90", color = "white", linewidth = 0.2) +
    geom_point(data = top1, aes(long, lat),
               size = 1.5, alpha = 0.6) +
    coord_fixed(1.3) +
    labs(title = "Heaviest 1 Percent - Locations",
         subtitle = "Basemap requires the maps package",
         x = "Longitude", y = "Latitude")
} else {
  # Fallback: plot points without a basemap so it never knits blank
  ggplot(top1, aes(long, lat)) +
    geom_point(size = 1.5, alpha = 0.6) +
    coord_fixed(1.3) +
    labs(title = "Heaviest 1 Percent - Locations",
         subtitle = "Basemap not installed. Showing points only.",
         x = "Longitude", y = "Latitude")
}
```

**Plain explanation:**  
These dots show where the biggest meteorites in the dataset were found. You will notice a lot in desert and open areas. That fits basic common sense. Big dark rocks are easier to see and keep their shape better on bare ground.

# Conclusion

Main takeaways in simple words:

- We record more meteorites per year now than we did long ago.
- The typical mass is smaller in recent decades, which fits with finding more small pieces.
- A few types are common almost every decade. Ordinary chondrites are the most common.
- The biggest pieces show up in many places, but deserts and open land make them easier to spot.

# Reproducibility

All results come directly from the code chunks. Knit this file and you should see the same numbers and figures.  
Keep "meteorite-landings.csv" in the same folder as this Rmd.

```{r session-info}
sessionInfo()
```
